Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-11-12 paper-new/concrete-relational-model.tex
some small writing tweaks to big-step term semantics and relational adequacy sections

Correspondingly, we define a version of the lock-step error ordering and weak
bisimilarity relation on $\delay(\mathbb{N} + {\mho})$ 
(see Section \ref{sec:relations-on-delay-monad} in the appendix for the definitions).
% \max{do these intermediate details matter for the main body of the paper?}
%
%
By adapting Theorem 4.3 of \cite{kristensen-mogelberg-vezzosi2022} to the
setting of inductively-defined relations, we can show that both the global
lock-step error ordering and the global weak bisimilarity admit coinductive
definitions. In particular, modulo the above isomorphism between $\li^{gl} X$
and $\delay(\mathbb{N} + {\mho})$, the global version of the lock-step error
ordering is equivalent to the lock-step error ordering on $\delay(\mathbb{N} +
{\mho})$, and the global version of weak bisimilarity is equivalent to weak
bisimilarity on $\delay(\mathbb{N} + {\mho})$.
%
This implies that the global denotation of the term precision semantics for
$\li^{gl} \mathbb{N}$ agrees with the corresponding relation $\ltsqbisim{}{}^\text{Delay}$ for $\delay(\mathbb{N} + {\mho})$, i.e.,
the closure of its lock-step ordering under weak bisimilarity on both sides.
%
Then to obtain the desired result, it suffices to show being related in
$\ltsqbisim{}{}^\text{Delay}$ implies that the corresponding partial elements of
$\mathbb{N} + {\mho}$ are related in the expected manner. That is, if $d
\ltsqbisim{}{}^\text{Delay} d'$, then $p(d) \ltdyn^{Partial} p(d')$ where $p$ is
a partial function from delays into $\mathbb{N} + {\mho}$ defined by existential
quantification over the number of steps $i$ in which $d$ terminates, and the
relation $\ltdyn^\text{Partial}$ is defined such that $\mho$ is the least
element, $\bot \ltdyn^\text{Partial} \bot$, and $n \ltdyn^\text{Partial} n$.
This implication follows from the definition of $\ltsqbisim{}{}^\text{Delay}$.

>>>>

Correspondingly, we define a version of the lock-step error ordering and weak
bisimilarity relation on $\delay(\mathbb{N} + {\mho})$ 
(see Section \ref{sec:relations-on-delay-monad} in the appendix for the definitions).
% \max{do these intermediate details matter for the main body of the paper?}
%
%
By adapting Theorem 4.3 of \cite{kristensen-mogelberg-vezzosi2022} to the
setting of inductively-defined relations, we can show that both the global
lock-step error ordering and the global weak bisimilarity admit coinductive
definitions. In particular, modulo the above isomorphism between $\li^{gl} X$
and $\delay(\mathbb{N} + {\mho})$, the global version of the lock-step error
ordering is equivalent to the lock-step error ordering on $\delay(\mathbb{N} +
{\mho})$, and the global version of weak bisimilarity is equivalent to weak
bisimilarity on $\delay(\mathbb{N} + {\mho})$.
%
This implies that the global denotation of the term precision semantics for
$\li^{gl} \mathbb{N}$ agrees with the corresponding relation $\ltsqbisim{}{}^\text{Delay}$ for $\delay(\mathbb{N} + {\mho})$, i.e.,
the closure of its lock-step ordering under weak bisimilarity on both sides.
%
Then to obtain the desired result, it suffices to show that being related in
$\ltsqbisim{}{}^\text{Delay}$ implies that the corresponding partial elements of
$\mathbb{N} + {\mho}$ are related in the expected manner. Specifically, recall
from Section \ref{sec:big-step-term-semantics} that $-\Downarrow^\text{Delay}$
is a partial function from delays to $\mathbb{N} + {\mho}$ defined by
existential quantification over the number of steps in which the argument terminates.
We want to show that if $d \ltsqbisim{}{}^\text{Delay} d'$, then 
%
$(d \Downarrow^\text{Delay}) \ltdyn^\text{Partial}\, (d' \Downarrow^\text{Delay})$,
%
where the relation $\ltdyn^\text{Partial}$ is defined such that $\mho$ is the
least element, $\bot \ltdyn^\text{Partial} \bot$, and $n \ltdyn^\text{Partial}
n$. This implication follows from the definition of
$\ltsqbisim{}{}^\text{Delay}$.

### 2024-11-12 paper-new/concrete-term-model.tex
small fixes to cast semantics figure
some small writing tweaks to big-step term semantics and relational adequacy sections

% TODO should we include function casts?
%% \begin{figure*}
%%   \begin{align*}
%%     \sem{\nat} &= \Nat \\
%%     \sem{\dyn} &= D \\
%%     \sem{A \times A'} &= \sem{A} \times \sem{A'} \\ 
%%     \sem{A \ra A'} &= \sem{A} \To \li \sem{A'} \\
%%   \end{align*}
%%   \caption{Interpretation of types}
%%   \label{fig:type-interpretation}
%% \end{figure*}
\begin{figure*}
  \begin{minipage}{0.3\textwidth}
    \begin{footnotesize}
  \begin{align*}
    % upcasts
    \sem{\upc{r(A)}} &= \id_{\sem{A}} \\
    \sem{\upc{(c \comp c')}} &= \sem{\upc{c'}} \circ \sem{\upc{c}} \\
    \sem{\upc{(c_i \ra c_o)}} &= (\sem{\dnc {c_i}} \tok \li\sem{A_o'})\circ (\sem{A_i} \tok U\li(\sem{\upc {c_o}}))\\
    \sem{\upc{(c_1 \times c_2)}} &= \sem{\upc{c_1}} \times \sem{\upc{c_2}}\\
    \sem{\upc{\inat}} &= \inat\\
    \sem{\upc{\itimes}} &= \itimes\\
    \sem{\upc{\iarr}} &= \iarr \circ \nxt\\
    % \sem{\zro}         &= \lambda \gamma . 0 \\
    % \sem{\suc\, V}     &= \lambda \gamma . (\sem{V}\, \gamma) + 1 \\
    % \sem{x \in \Gamma} &= \lambda \gamma . \gamma(x) \\
    % \sem{\lda{x}{M}}   &= \lambda \gamma . \lambda a . \sem{M}\, (*,\, (\gamma , a))  \\
    % \sem{V_f\, V_x}    &= \lambda \gamma . {({(\sem{V_f}\, \gamma)} \, {(\sem{V_x}\, \gamma)})} \\
    % \sem{\err_B}       &= \lambda \gamma . \mho \\
    %
    % downcasts
    \sem{\dnc{r(A)}} &= \id_{\li \sem{A}} \\
    \sem{\dnc{(c \comp c')}} &= \sem{\dnc{c}} \circ \sem{\dnc{c'}} \\
    \sem{\dnc{(c_i \ra c_o)}} &= U(\sem{\upc {c_i}} \to \sem{\dnc {c_o}})\\
    \sem{\dnc{(c_1 \times c_2)}} &= (\dnc{c_1} \timesk A_2) \circ (A_1' \timesk \dnc{c_2}) \\
    \sem{\dnc{\inat}} &= (
      \lambda V_d . \text{case $({V_d})$ of}
      \{ \inat\,n \to \eta\, n
         \alt \text{otherwise} \to \mho \})^\dagger \\
    \sem{\dnc{\itimes}} &= (
      \lambda V_d . \text{case $({V_d})$ of}
      \{ \itimes(d_1, d_2) \to \eta\, (d_1, d_2)
         \alt \text{otherwise} \to \mho \})^\dagger \\
    \sem{\dnc{\iarr}} &= (
      \lambda V_d . \text{case $({V_d})$ of}
      \{ \iarr \tilde{f} \to \theta_t\, (\eta (\tilde{f}_t))
         \alt \text{otherwise} \to \mho \})^\dagger \\
    % \sem{\ret\, V}       &= \lambda \gamma . \eta\, \sem{V} \\
    % \sem{\bind{x}{M}{N}} &= \lambda \delta . \ext {(\lambda x . \sem{N}\, (\delta, x))} {\sem{M}\, \delta} \\
  \end{align*}
    \end{footnotesize}
  \end{minipage}
  \caption{Semantics of casts}
  \label{fig:term-semantics}
\end{figure*}

@@

Consider the functor $F(X) = \mathbb{N} + {\mho} + X$. By the definition of the
guarded lift, we have that
% 
$\li^k \mathbb{N} 
  \cong \mathbb{N} + {\mho} + \later^k (\li \mathbb{N}) 
  \cong F(\later \li^k \mathbb{N})$.
%
Furthermore, it can be shown that the functor $F$ satisfies the technical
condition in the statement of the above-mentioned theorem. Thus, the theorem
implies that the type $\li^{gl} \mathbb{N} := \forall k. \li^k \mathbb{N}$ is a
final $F$-coalgebra. In fact, $\li^{gl}$ is isomorphic to a more familiar coninductive type.
%
Given a type $Y$, Capretta's \emph{delay monad} \cite{lmcs:2265},
which we write as $\text{Delay}(Y)$, is defined as the coinductive type
generated by $\tnow : Y \to \delay(Y)$ and $\tlater : \delay(Y) \to \delay(X)$.
We have that $\delay (\mathbb{N} + {\mho})$ is a final coalgebra for the functor
$F$ defined above. This means $\li^{gl} \mathbb{N}$ is isomorphic to
$\delay(\mathbb{N} + {\mho})$. We write $\alpha$ to denote this isomorphism.

Given $d : \delay(\mathbb{N} + {\mho})$, we define a notion of termination in
$i$ steps with result $n_? \in \mathbb{N} + {\mho}$, written $d \da^i n_?$.
We define $d \da^i n_?$ inductively by the rules
$(\tnow\, n_?) \da^0 n_?$ and $(\tlater\, d) \da^{i+1} n_?$ if $d \da^i n_?$.
From this definition, it is clear that if there is an $i$
such that $d \da^i n_?$, then this $i$ is unique. 
% Furthermore, from the proof that
% $d$ terminates in $i$ steps, we can extract the result with which $d$
% terminates. This is defined by a straightforward induction on the proof that $d \da^i$.
Using this definition of termination, we define a partial function 
%
$f \colon \delay (\mathbb{N} + {\mho}) \rightharpoonup \mathbb{N} + {\mho}$
%
by existentially quantifying over $i$ for which $d \da^i n_?$. That is, we
define $f(d) := n_?$ if $d \da^i n_?$ for some (necessarily unique) $i$, and $f$
is undefined otherwise.

Now consider the global version of the term semantics, $\sem{\cdot}^{gl} \colon
\{M \,|\, \cdot \vdash M : \nat \} \to \li^{gl} \mathbb{N}$, defined by
$\sem{M}^{gl} := \lambda (k : \Clock) .\sem{M}$. Composing this with the
isomorphism $\alpha$ and the partial function $f$ yields the desired partial
function
%
$-\Downarrow \colon \{M \,|\, \cdot \vdash M : \nat \} \rightharpoonup \mathbb{N} + {\mho}$.
%
It is then straightforward to see that this big-step semantics respects the
equational theory and that the denotations of the diverging term $\Omega$, the
error term $\mho$, and a syntactic natural number value $n$ are all distinct.

>>>>

% TODO should we include function casts?
%% \begin{figure*}
%%   \begin{align*}
%%     \sem{\nat} &= \Nat \\
%%     \sem{\dyn} &= D \\
%%     \sem{A \times A'} &= \sem{A} \times \sem{A'} \\ 
%%     \sem{A \ra A'} &= \sem{A} \To \li \sem{A'} \\
%%   \end{align*}
%%   \caption{Interpretation of types}
%%   \label{fig:type-interpretation}
%% \end{figure*}
\begin{figure*}
  \begin{minipage}{0.3\textwidth}
    \begin{footnotesize}
  \begin{align*}
    % upcasts
    \sem{\upc{r(A)}} &= \id_{\sem{A}} \\
    \sem{\upc{(c \comp c')}} &= \sem{\upc{c'}} \circ \sem{\upc{c}} \\
    \sem{\upc{(c_i \ra c_o)}} &= (\sem{\dnc {c_i}} \tok \li\sem{A_o'})\circ (\sem{A_i} \tok U\li(\sem{\upc {c_o}}))\\
    \sem{\upc{(c_1 \times c_2)}} &= \sem{\upc{c_1}} \times \sem{\upc{c_2}}\\
    \sem{\upc{\inat}} &= \inat\\
    \sem{\upc{\itimes}} &= \itimes\\
    \sem{\upc{\iarr}} &= \iarr \circ \nxt\\
    % \sem{\zro}         &= \lambda \gamma . 0 \\
    % \sem{\suc\, V}     &= \lambda \gamma . (\sem{V}\, \gamma) + 1 \\
    % \sem{x \in \Gamma} &= \lambda \gamma . \gamma(x) \\
    % \sem{\lda{x}{M}}   &= \lambda \gamma . \lambda a . \sem{M}\, (*,\, (\gamma , a))  \\
    % \sem{V_f\, V_x}    &= \lambda \gamma . {({(\sem{V_f}\, \gamma)} \, {(\sem{V_x}\, \gamma)})} \\
    % \sem{\err_B}       &= \lambda \gamma . \mho \\
    %
    % downcasts
    \sem{\dnc{r(A)}} &= \id_{\li \sem{A}} \\
    \sem{\dnc{(c \comp c')}} &= \sem{\dnc{c}} \circ \sem{\dnc{c'}} \\
    \sem{\dnc{(c_i \ra c_o)}} &= U(\sem{\upc {c_i}} \to \sem{\dnc {c_o}})\\
    \sem{\dnc{(c_1 \times c_2)}} &= (\sem{\dnc{c_1}} \timesk \sem{A_2}) \circ (\sem{A_1'} \timesk \sem{\dnc{c_2}}) \\
    \sem{\dnc{\inat}} &= (
      \lambda V_d . \text{case $({V_d})$ of }
      \{ \inat\,n \to \eta\, n
         \alt \text{otherwise} \to \mho \})^\dagger \\
    \sem{\dnc{\itimes}} &= (
      \lambda V_d . \text{case $({V_d})$ of }
      \{ \itimes(d_1, d_2) \to \eta\, (d_1, d_2)
         \alt \text{otherwise} \to \mho \})^\dagger \\
    \sem{\dnc{\iarr}} &= (
      \lambda V_d . \text{case $({V_d})$ of }
      \{ \iarr \tilde{f} \to \theta_t\, (\eta (\tilde{f}_t))
         \alt \text{otherwise} \to \mho \})^\dagger \\
    % \sem{\ret\, V}       &= \lambda \gamma . \eta\, \sem{V} \\
    % \sem{\bind{x}{M}{N}} &= \lambda \delta . \ext {(\lambda x . \sem{N}\, (\delta, x))} {\sem{M}\, \delta} \\
  \end{align*}
    \end{footnotesize}
  \end{minipage}
  \caption{Semantics of casts}
  \label{fig:term-semantics}
\end{figure*}

@@

Consider the functor $F(X) = \mathbb{N} + {\mho} + X$. By the definition of the
guarded lift, we have that
% 
$\li^k \mathbb{N} 
  \cong \mathbb{N} + {\mho} + \later^k (\li \mathbb{N}) 
  \cong F(\later \li^k \mathbb{N})$.
%
Furthermore, it can be shown that the functor $F$ satisfies the technical
condition in the statement of the above-mentioned theorem. Thus, the theorem
implies that the type $\li^{gl} \mathbb{N} := \forall k. \li^k \mathbb{N}$ is a
final $F$-coalgebra. In fact, $\li^{gl}$ is isomorphic to a more familiar coninductive type.
%
Given a type $Y$, Capretta's \emph{delay monad} \cite{lmcs:2265},
which we write as $\text{Delay}(Y)$, is defined as the coinductive type
generated by $\tnow : Y \to \delay(Y)$ and $\tlater : \delay(Y) \to \delay(Y)$.
We have that $\delay (\mathbb{N} + {\mho})$ is a final coalgebra for the functor
$F$ defined above. This means $\li^{gl} \mathbb{N}$ is isomorphic to
$\delay(\mathbb{N} + {\mho})$. We write $\alpha$ to denote this isomorphism.

Given $d : \delay(\mathbb{N} + {\mho})$, we define a notion of termination in
$i$ steps with result $n_? \in \mathbb{N} + {\mho}$, written $d \da^i n_?$.
We define $d \da^i n_?$ inductively by the rules
$(\tnow\, n_?) \da^0 n_?$ and $(\tlater\, d) \da^{i+1} n_?$ if $d \da^i n_?$.
From this definition, it is clear that if there is an $i$
such that $d \da^i n_?$, then this $i$ is unique. 
% Furthermore, from the proof that
% $d$ terminates in $i$ steps, we can extract the result with which $d$
% terminates. This is defined by a straightforward induction on the proof that $d \da^i$.
Using this definition of termination, we define a partial function 
%
$-\Downarrow^\text{Delay} \colon \delay (\mathbb{N} + {\mho}) \rightharpoonup \mathbb{N} + {\mho}$
%
by existentially quantifying over $i$ for which $d \da^i n_?$. That is, we
define $d \Downarrow^\text{Delay} \,= n_?$ if $d \da^i n_?$ for some (necessarily
unique) $i$, and $d \Downarrow^\text{Delay}$ is undefined otherwise.

Now consider the global version of the term semantics, $\sem{\cdot}^{gl} \colon
\{M \,|\, \cdot \vdash M : \nat \} \to \li^{gl} \mathbb{N}$, defined by
$\sem{M}^{gl} := \lambda (k : \Clock) .\sem{M}$. Composing this with the
isomorphism $\alpha : \li^{gl} \mathbb{N} \cong \delay(\mathbb{N} + {\mho})$ and
the partial function $-\Downarrow^\text{Delay}$ yields the desired partial
function
%
$-\Downarrow \colon \{M \,|\, \cdot \vdash M : \nat \} \rightharpoonup \mathbb{N} + {\mho}$.
%
It is then straightforward to see that this big-step semantics respects the
equational theory and that the denotations of the diverging term $\Omega$, the
error term $\mho$, and a syntactic natural number value $n$ are all distinct.
